---
title: "Gender Pay Gap Methods"
format: html
execute: 
  warning: false
  message: true
  echo: true
#context: setup

---

## General background
- Gender Pay Gap calculation is a crucial requirements for most companies, given new laws.

## What is commonly reported
- Normally companies are interested in reporting:
    - Unadjusted GPG
    - Adjusted GPG

There is consensus around UGPG but not around AGPG

## Issues with modeling adjusted gender pay gap
- Comparability (what goes in the model)
- Modeling (what methodologies)
- Reporting (estimates vs uncertainties)

## Methodologies described
- Oaxaca Blinder
- IPW
- Matching


#First do some basic descriptives
#then, introduce the standard approach
#then, introduce the issue of matching by ipw
#finally, introduce matching by matching options and marginal effects

```{r}
df = read.csv("Glassdoor Gender Pay Gap.csv")
head(df)
```

# Data Exploration
Next we are just going to look at some basic descriptives for some of the variables by gender. 

```{r, echo = FALSE}
#| fig-width: 12
#| fig-height: 10
# Load necessary libraries
library(psych)
library(GGally)
library(oaxaca)


# Remove any rows with missing data
df_clean <- na.omit(df)

# Generate the ggpairs plot with boxplots and correlations
pm <- ggpairs(
  df_clean[, c(3,4,7,5)], # Include the desired columns
  mapping = aes(colour = as.factor(df_clean$Gender)), # Specify the color mapping here
  title = "Plots and Correlations for GlassDoor Data by Male and Female" # Optional: Add title
  )


# Display the plot
print(pm)

```

What we can observe is that there seems to be some difference in terms of Age, performance evaluation, seniority and education.

# Unadjusted Pay Gap

The first step in analyzing the GPG would be the average difference in pay between man and women, also called the Unadjusted Gender Pay gap.
In our case we will be following the EU directive on this metric https://ec.europa.eu/eurostat/cache/metadata/en/earn_grgpg2_esms.htm#:~:text=according%20to%20the%20following%20definition,paid%20employees%5D%20expressed%20in%20%25.
which is GPG = [(average gross hourly earnings of male paid employees - average gross hourly earnings of female paid employees) / average gross hourly earnings of male paid employees] expressed in %

```{r}
avg_pay_men = mean(df[df['Gender'] == 'Male', 'BasePay'])
avg_pay_women = mean(df[df['Gender'] == 'Female', 'BasePay'])
GPG = round(((avg_pay_men-avg_pay_women)/avg_pay_men)*100,2)

paste0("The Unadjusted Gender Pay Gap is = ", GPG, "%")

```

The resulting gender pay gap is 8.65%, indicating that women, on average, earn 8.65% less than men. 

# Adjusted GPG
## Oaxaca Blinder
```{r}
library(oaxaca)
library(psych)
df_clean$BasePay_log = log(df_clean$BasePay)
df_clean$Gender_fac = ifelse(df_clean$Gender == "Male", 0, 1)
oaxaca.results = oaxaca(BasePay_log ~ Age + PerfEval + Education + Seniority + Dept | Gender_fac, data = df_clean)

AGPG = round(((oaxaca.results$y$y.A - oaxaca.results$y$y.B)/oaxaca.results$y$y.A)*100,2)

paste0("The Adjusted Gender Pay Gap is = ", AGPG, "%")
```